\chapter{Discussion and Conclusion}
This Master's thesis investigated the reduction in parameter count for \gls{DiT}s to enable more efficient image generation and facilitate their deployment in resource-constrained environments, particularly as new models become increasingly large and computationally intensive. An in-depth analysis of the compression framework was conducted on PixArt-$\Sigma$ (Sec. \ref{sec:PixArtExp}), evaluating several design choices and transferring the optimal settings to FLUX.1 {[dev]} (Sec. \ref{sec:ExpFlux}). Surprisingly, the analysis revealed that low-rank approximation via truncated \gls{SVD} is a highly effective method for compressing the projection matrices in individual Transformer blocks while maintaining strong performance. This contrasts with earlier studies on \gls{LLM}s, which demonstrated that ordinary low-rank approximation led to significant quality loss and more advanced methods were required. Therefore, an importance-aware rank-adaptive progressive compression approach, \textit{SVDtrunc}, was developed, establishing a new quality-parameter frontier for FLUX.1 {[dev]}.


In previous structural pruning methods, model components were either removed entirely or replaced by a smaller module. In contrast, \gls{SVD}trunc leaves the overall architecture intact by performing a low-rank approximation of the weight matrices. Sec. \ref{sec:CompressionStrategyPixart} and \ref{sec:trainingFreeCompression} underline that low-rank approximation provides a robust initialization for compressed models, leading to superior results compared to complete block pruning. Particularly, for the training-free setting (Sec. \ref{sec:trainingFreeCompression}), \gls{SVD}trunc outperforms EcoDiff \cite{EcoDiff} and the block pruning strategy for models smaller than 90\% of their original size. Especially, when retaining only 70\% of the original parameters, EcoDiff and block pruning collapse completely, while \gls{SVD}trunc still generates viable images, even though a degradation in quality is observed. This suggests that the redundancies in \gls{DiT}s are primarily located within individual Transformer blocks rather than across them and that each block fulfills its own functional task. Thus, by removing an entire block, the neighboring blocks must compensate for this loss, potentially limiting their capacity to perform their own functions. 

The analysis further revealed that an importance-aware block selection strategy outperforms random or uniform block selection (Sec. \ref{sec:FluxBlockImportanceAnalysis}). Moreover, instead of applying an equal compression rate across all selected blocks, a decreasing linear compression schedule was employed, compressing less important blocks more aggressively. This adaptive distribution of discarded parameters over the selected blocks demonstrated superior performance compared to a fixed-rate baseline (Sec. \ref{sec:CompressionSchedulingPixart}). These findings suggest that individual blocks process tasks of varying complexity. Consequently, they exhibit distinct levels of resilience to compression before experiencing measurable performance degradation.

The effectiveness of \gls{SVD}trunc was demonstrated on FLUX.1 {[dev]} (Sec. \ref{sec:FluxMainResults}).  \gls{SVD}trunc-m, which is compressed to 68\% of the original model size of FLUX.1 {[dev]}, performs nearly as well as FLUX.1 {[dev]}, only an average degradation of 0.75\% across three benchmarks (HPSv2, GenEval and DPG) is observed. As the model size was reduced further, the overall benchmark performance dropped gradually and predictably, such that even \gls{SVD}trunc-s, containing 57\% of the original number of parameters, outperforms competing methods of significantly larger model sizes. 


The investigation of FLUX.1 {[dev]} uncovered two distinct phenomena. First, a gradual domain shift was occasionally observed across different compression ratios, for instance from anime to realistic styles. Second, a loss of memorization regarding highly distinct concepts, such as "Bugs Bunny", became apparent while reducing model size. Notably, in both scenarios, even strongly compressed models (e.g. up to 39\% of the original size) continue to generate plausible, aesthetically pleasing images. This demonstrates that truncated \gls{SVD} is very effective in preserving parameters necessary for general image generation. However, fine details such as style specific nuances or highly specialized concepts appear to be stored in low-magnitude singular values that are removed and cannot be recovered. Consequently, while \gls{SVD}trunc is ideal for general-purpose deployment where memory efficiency is paramount, specialized applications requiring high fidelity to specific artistic styles or intellectual properties might require more targeted preservation strategies.

%Consequently, \gls{SVD}trunc acts as a semantic filter that prioritizes structural coherence over detail.

Despite the superior benchmark performance and the comparable reduction in \gls{VRAM} consumption relative to competing methods, a notable limitation of \gls{SVD}trunc is that the inference time decreases only marginally due to the increased number of sequential matrix operations. For this reason, to achieve near real-time applications, this approach must be coupled with a reduction in inference steps. Therefore, it is imperative that \gls{SVD}trunc is compatible with step-distillation methods to enable the development of compact, fast, and high-quality models. Sec. \ref{sec:FluxSchnellMainResults} demonstrates that for compression ratios up to 30\%, the performance degradation of FLUX.1 {[schnell]} is minor ($R=2.72\%$). However, when compressing FLUX.1 {[schnell]} to 50\% of its original size, a significant performance drop occurs ($R=10.69\%$). The resulting images appear washed out and exhibit lower quality. This indicates that step-distillation methods may contain fewer parameter redundancies and therefore cannot be compressed as aggressively as FLUX.1 {[dev]}. Furthermore, the washed-out appearance closely resembles images generated by ordinary \gls{DiT}s using an insufficient number of denoising steps. Therefore, another reason for the performance drop might be that the model's few-step generation capabilities acquired through step-distillation are compromised. Consequently, future work could investigate whether it is more effective to reverse the order by compressing the model first and subsequently applying step distillation. Alternatively, exploring a joint schedule integrating iterative compression and step-distillation training could be beneficial to better maintain the model's performance.


Future research should also explore ``importance-aware`` truncation strategies, where singular values are preserved not based on their magnitude alone but on their contribution to specific, highly distinctive concepts or styles. A potentially promising approach is \gls{GRASP}, which precisely demonstrated in the training-free setting that selecting singular values based on task importance rather than magnitude enables the generation of images with greater detail (Sec. \ref{sec:GRASPForDiTs}). Therefore, investigating whether this advantage persists after subsequent fine-tuning represents a possible next step.

Beyond importance-aware singular value selection, another avenue that future research must tackle is the discrepancy between theoretical FLOPs and practical inference time. To reduce the actual latency, the development of hardware-optimized CUDA kernels specifically designed for low-rank matrix multiplication could be explored. Furthermore, examining whether \gls{SVD}trunc can be combined with low-bit quantization to further push the boundaries of deployment on consumer-grade hardware with even stricter \gls{VRAM} limitations.