\pagenumbering{arabic}
\chapter{Introduction}
Over the past few years, generative artificial intelligence (\gls{GenAI})  has gained immense importance and relevance in society, as well as in business and politics. In addition to large language models (\gls{LLM}s) \cite{Claude,GPT4,Gemini,Lama}, which have brought \gls{GenAI} into the mainstream, generative image models \cite{Flux,Pixart_s,baldridge2024imagen,brooks2024video,SD21,SDXL} are receiving increasing attention. The continuous surge in the number of image generator models and simplified access present both opportunities and risks. On the one hand, they open up a range of new possibilities in areas such as design and marketing as well as gaming and entertainment. On the other hand, they enable the generation of highly realistic synthetic images and videos, often referred to as deep fakes, raising severe ethical and privacy concerns. Consequently, policymakers are striving to establish rule-based guidelines for their use (European AI Act \cite{eu_ai_act_2024}). 

The advent of deep learning based image generators is often marked by the introduction of variational autoencoders (\gls{VAE}s) in 2013 \cite{VAE}, as they demonstrated that neural networks can learn complex data distribution and generate novel samples. Subsequently, generative adversarial networks (\gls{GAN}s) \cite{GAN,StyleGAN_3,StyleGAN_T} became the dominant model type in the image generation field as they produce images of significantly higher quality than \gls{VAE}s. However, \gls{GAN}s are notoriously difficult to train, as they often exhibit mode collapse and suffer from exploding or vanishing gradients \cite{ModeCollapse1,ImprovedGANTraining, PacGAN}.

 In recent years, \gls{GAN}s were largely surpassed by diffusion models \cite{DDPM} deploying an iterative image generation denoising process. The breakthrough of diffusion models as the new state-of-the-art (\gls{SOTA}) in image generation occurred mainly in 2022 when the latent diffusion models, most notably Stable Diffusion \cite{LDM}, were released. At the time, they excelled in image quality and image generation diversity, but came at the cost of high inference time due to the iterative nature of their generation process. The transition from pixel space to latent space massively reduced the computational demands of the models and enabled the generation of high-resolution images.
 
  The first Stable Diffusion models \cite{SD21,SDXL} leveraged a U-Net architecture \cite{Unet} and already required a huge amount of training data and computational resources. Their image generation process is typically controlled by textual prompts integrated via cross-attention mechanisms \cite{AttentionIsAllYouNeed}. To expand these capabilities, several new control and fine-tuning methods like ControlNet \cite{ControlNet} and DreamBooth \cite{ruiz2023dreamboothfinetuningtexttoimage} were developed to enable precise spatial conditioning and personalization. Since the U-Net-based SD models already included transformer layers \cite{AttentionIsAllYouNeed} for text-conditioning, the next logical evolution was to transition to purely transformer-based models known as diffusion transformers (\gls{DiT}s) \cite{DiT}, which scaled significantly in parameter size and required even more training data. In addition, a new formulation, so-called flow-matching \cite{FlowMatching,RectifiedFlowMatching}, became predominant due to improved training stability and sampling efficiency which is leveraged by the current SOTA image generators like FLUX.1 [dev] \cite{Flux} containing 11.9 billion parameters.
  
  Possible areas of application for diffusion and flow models include edge devices such as mobile phones \cite{li2023snapfusion,chen2023speed,zhao2024mobilediffusion}. In these cases, it is advantageous to run the models directly on the edge device in order to avoid server costs, to make the associated functionality available offline and to protect the privacy of users so that, for instance, images do not have to be uploaded to external servers for processing.
  However, current flow-based models like FLUX.1 [dev] face two distinct challenges for real-time applications on edge devices. On the one hand, as mentioned before, they suffer from a long inference time due to the sequential application of the model. On the other hand, these models also require a substantial amount of \gls{VRAM} during inference, e.g. FLUX.1 [dev] utilizes about 32GB. Therefore, reducing the computational requirements of \gls{DiT}s is essential for their deployment.
 
  
  Besides the strict constraints on edge devices, lowering the computational demands for \gls{DiT}s plays also a pivotal role in democratizing these models and making them more environmentally friendly, as they can run on consumer-grade hardware and do not require power-intensive data centres, thereby reducing their carbon footprint. 

  To overcome these fundamental limitations, recent research has focused on mitigating both the latency and memory bottlenecks. Regarding the first challenge, the main research stream for speeding up the inference time is to enable the model to generate high-quality images with fewer inference steps. While at the beginning, diffusion models required up to 1000 inference steps to generate one single image, using more advanced \gls{ODE}-solvers \cite{DDIM,lu2025dpm} for the denoising trajectory has led to a substantial reduction in sampling time.
  
   Building on these algorithmic improvements, another major line of research involves step- and guidance distillation. Unlike solver-based acceleration, distillation methods transfer the capabilities of a complex teacher model to a more efficient student model. In step distillation, the same model is often used as both the teacher and student model, and the student is trained to replicate the teacher's output quality in as few as four or two, sometimes even only a single inference step \cite{ADD,LADD,ProgressiveDistillation,ConsistencyModel,LatentConsistencyModel,DistillationMatching}. A prominent example of this technique is the transition from FLUX.1 [dev] to FLUX.1 [schnell], reducing the required inference steps to only four.
  
   Complementary to reducing the step count, guidance distillation aims to incorporate the concept of classifier free guidance (\gls{CFG}), which normally requires the model to be applied twice per inference step, into the model itself, thereby saving half of the model evaluations \cite{GuidedDistillation}. This procedure was utilized to distill Flux.1-pro into FLUX.1 [dev].
   
  While step- and guidance distillation effectively reduce the latency (inference time), they do not necessarily address the second bottleneck, namely high \gls{VRAM} consumption. Proposed methods for \gls{VRAM} reduction range from quantization \cite{han2015deep,FluxBit} and more efficient implementation of model components, e.g. Flash Attention \cite{FlashAttention}, to structured or unstructured model pruning \cite{Bksdm,Laptop,Koala,FastFlux,flux1-lite,flux_mini_2024,EcoDiff,PPCL,HierarchicalPrune}. While quantization compresses the model size by reducing the precision of the model parameters, pruning directly addresses the phenomenon that modern \gls{DiT}s exhibit high redundancy in and between the individual Transformer blocks. Current methods can be categorized into two main approach, where model parts are either completely removed \cite{flux1-lite,Bksdm} or are replaced by smaller modules e.g. a linear layer \cite{FastFlux}. For this reason, identification followed by compression of redundant model parts is a promising orthogonal approach to the reduction of inference steps for further efficiency gains and is therefore the main focus of this Master's thesis.
  
   The first part of this Master's thesis systematically examines how the number of parameters in \gls{DiT}s can be effectively reduced while minimizing degradation of image generation quality. PixArt-$\Sigma$ \cite{Pixart_s} is utilized as a representative baseline enabling the execution of numerous experiments due to its significantly smaller parameter count compared to FLUX.1 [dev]. Thereafter, the identified compression framework is applied to FLUX.1 [dev] and FLUX.1 [schnell] to demonstrate its effectiveness in the high-compression regime and for step-distilled models. 
   
The analysis of compressing PixArt-$\Sigma$ spans three dimensions. The first dimension deals with the identification of redundant model components. The second dimension examines how redundant model components can be compressed most effectively with minimal loss of quality. And the third dimension investigates how the generative capacities of the \gls{DiT}s can be recovered optimally after compression through fine-tuning.

%	The analysis revealed that redundancies in \gls{DiT}s are primarily localized within individual Transformer blocks rather than distributed across them. Consequently, the complete pruning of a block forces adjacent blocks to compensate for the missing functionality, potentially reducing their capacity to perform their original tasks. Furthermore, the analysis demonstrated that applying low-rank approximation via truncated singular value decomposition (\gls{SVD}) of projection matrices within Transformer blocks is highly effective at reducing the parameter count in \gls{DiT}s, in stark contrast to \gls{LLM}s where naive truncated \gls{SVD} leads to a severe drop in performance \cite{hsu2022language,wang2025dobi,SVD_LLM_V2,wang2024svd}. Notably, \gls{DiT}s even maintain acceptable generation capabilities after \gls{SVD}-based compression without subsequent fine-tuning.
	
	
	Based on this analysis, the main contributions of this Master's thesis are the following: \textit{(1)} It is demonstrated that \gls{DiT}s are highly robust against low-rank approximation of their projection matrices, as increased compression results in a gradual decline in performance. This behavioral pattern fundamentally differs from \gls{LLM}s where naive truncated \gls{SVD} leads to a severe drop in performance \cite{hsu2022language,wang2025dobi,SVD_LLM_V2,wang2024svd}. The high tolerance of \gls{DiT}s becomes especially evident through the strong performance of \gls{SVD}-based compression even without subsequent fine-tuning. \textit{(2)} It is shown that redundancies in \gls{DiT}s are primarily localized within Transformer blocks, specifically distributed across projection matrices, and not across multiple Transformer blocks. \textit{(3)} An importance-aware, rank-adaptive progressive compression protocol on the block level is established that iteratively ranks blocks according to their task importance, compresses less important blocks more aggressively, and subsequently leverages knowledge distillation for fine-tuning. This protocol is termed \textit{SVDtrunc}. \textit{(4)} \gls{SVD}trunc sets a novel quality-parameter frontier for FLUX.1 [dev], outperforming all competing methods. 
	
   
   
   
   The remainder of this Master's thesis is structured as follows:  Chapter 2 provides a comprehensive overview of image generator models, emphasizing diffusion and flow models, alongside relevant metrics used for measuring performance. Building on this context, Chapter 3 contextualizes this work within the current literature by reviewing alternative model compression techniques. With the background established, Chapter 4 introduces the specific methodologies investigated in this work. Their effectiveness is then rigorously evaluated through experimental results in Chapter 5. Finally, Chapter 6 concludes the thesis with a comprehensive discussion of the main findings and their implications.
   
   %	Overall, an iterative parameter reduction framework, termed \textit{\gls{SVD}trunc}, proved to be the most efficient. It employes an importance-aware rank-adaptive progressive compression strategy via low-rank approximation, compressing less important blocks more aggressively. Subsequently, knowledge distillation is leveraged for recovering the model's generation capabilities. In conclusion, \gls{SVD}trunc establishes a new quality-parameter frontier for FLUX.1 [dev] compared to competing compression methods and it is demonstrated that the framework can also be applied to step-distilled models.
   
   %	The analysis revealed that redundancies in \gls{DiT}s are primarily localized within individual Transformer blocks rather than distributed across them. Consequently, the complete pruning of a block forces adjacent blocks to compensate for the missing functionality, potentially reducing their capacity to perform their original tasks. Furthermore, the analysis demonstrated that applying low-rank approximation via truncated singular value decomposition (\gls{SVD}) of projection matrices within Transformer blocks is highly effective at reducing the parameter count in \gls{DiT}s, in stark contrast to \gls{LLM}s where naive truncated \gls{SVD} leads to a severe drop in performance \cite{hsu2022language,wang2025dobi,SVD_LLM_V2,wang2024svd}. Notably, \gls{DiT}s even maintain acceptable generation capabilities after \gls{SVD}-based compression without subsequent fine-tuning.
   
%   
%   Specifically, the identification of architectural components with redundant parameters alongside the best pruning strategies, e.g., complete layer removal versus low-rank compression, and pruning schedules, e.g., one-shot versus progressive pruning are investigated. 
%   Furthermore, various post-compression retraining frameworks are evaluated to optimize performance recovery. The empirical results demonstrate that progressively compressing model blocks via low-rank approximation leads to the best results. Hereby, a linear distribution of compression ratios over transformer blocks meaning compressing the least important structures most aggressively and more critical structures less, is leveraged. 
%   
%   In the second part of this Master's thesis, the previously identified best pruning strategy is applied to the FLUX.1 [dev] model and compared with other existing compression methods.\\
%    Furthermore, a training-free compression strategy originally proposed for \gls{LLM}s, namely \gls{GRASP}, that is also based on singular value decomposition (\gls{SVD}) is transferred to FLUX.1 [dev] and tested in comparison with current alternative approach \cite{EcoDiff}. \\\\
%The remainder of this Master's thesis is structured as follows: Chapter two provides a comprehensive overview of image generator models, emphasizing diffusion and flow models, alongside relevant metrics used for measuring the performance. Building up on this context, chapter three contextualizes this work within the current literature by reviewing alternative model compression techniques. With the background established, chapter four introduces the specific methodologies investigated in this work. Their effectiveness is then rigorously evaluated through experimental results in chapter five. Finally, chapter six concludes the thesis with a comprehensive discussion of the main findings and their implications.
%  
%  
%  

%   Famous distillation frameworks are latent adversarial diffusion distillation \cite{ADD,LADD} which utilizes besides a teacher model a discriminator as additional learning signal which differentiate if an image was generated by the student model or is from the real dataset or latent consistency models \cite{ConsistencyModel,LatentConsistencyModel} which are designed to map images with different noise levels to their corresponding clean image in one step by enforcing the self-consistency property meaning that two noisy images lying on the same denoising path are always mapped to the same clean image.
