\chapter{Related Work}

Improvements in model performance are achieved not only through new algorithms and architectures, but also through the continuous scaling of model size. This scaling paradigm is facilitated due to the continuous advancements in computation hardware and the availability of internet scale training datasets. While massive architectures yields superior generative capabilities, there is a growing need to achieve comparable performance with smaller models. The hardware required for such large models is extremely expensive and power-intensive. The latter aspects leads to a substantial carbon footprint and causes environmental concerns. Furthermore, deploying these models is often infeasible in a resource-constrained setup, e.g. on edge devices. Consequently, a diverse body of literature has emerged around model compression techniques, aiming to minimize the parameter count and the memory footprint during inference and, ideally, also reducing the inference time. This chapter provides a comprehensive overview of various model compression methods, with a particular focus on methodologies adapted for \gls{DiT}s as well as methods to accelerate inference. Both are critical for edge device deployment.

 \subsubsection{Step Distillation}
Diffusion and flow models suffer from long inference times due to their iterative image generation process, which requires repeated evaluations of the model. Step distillation methods achieve a significant speed-up of the image generation process by reducing the number of required inference steps down to four or even a single step \cite{LADD,ADD,ProgressiveDistillation,DistillationMatching,ConsistencyModel,LatentConsistencyModel}. In step distillation frameworks, the teacher and the student model often share the same architecture. The teacher is leveraged to re-train the student to generate images utilizing fewer inference steps. To this end, several different approaches have been proposed.

One foundational approach is progressive distillation \cite{ProgressiveDistillation}, in which the number of inference steps is iteratively halved and the previous student model is utilized as the new teacher model in the next iteration. Consistency models \cite{ConsistencyModel,LatentConsistencyModel} even achieve a one-step image generation process by enforcing the self-consistency property which dictates that all points on a given diffusion trajectory must be mapped to the same clean image. 
Alternatively, rather than strictly following the teacher's trajectory, recent methods rely on adversarial or statistical matching. Latent adversarial diffusion distillation (\gls{LADD}) \cite{ADD,LADD} incorporates a discriminator in a GAN-like training setup to provide an additional learning signal. Similarly, moving away from exact sample regression,  distribution matching distillation \cite{DistillationMatching} focuses on matching the entire image distribution. Specifically, it optimizes the student model to produce samples that collectively match the statistical properties of the real data, and not just the output of the teacher on specific noise-image pairs.

Importantly, these step distillation approaches are orthogonal to structural model compression. While this Master's thesis focuses on architectural model reduction via truncated \gls{SVD}, step distillation can be seamlessly integrated with compressed models to achieve even higher inference efficiency which is demonstrated on FLUX.1 [schnell] (Sec. \ref{sec:FluxSchnellMainResults}).

\subsubsection{Quantization}
Quantization is a widely adopted technique to reduce model size and speed up the inference process \cite{cai2020zeroq,jacob2018quantization,han2015deep}. In quantization, the high-precision weight values (e.g. FP32) are mapped to discrete or lower-precision values (e.g. INT8) thereby significantly reducing the memory footprint required to store the weights. Quantization methods can be broadly categorized into two main approaches.

First, post-training quantization (\gls{PTQ}) \cite{PTQD,jacob2018quantization, nagel2020up} encompasses all methods that operate on fully trained models. They typically rely on a small calibration dataset to determine the scaling and zero-point parameters for mapping the high-precision values to their discrete low-precision counterparts. These methods are computationally highly efficient, as no full retraining of the complete, potentially very large model is necessary. 

Second, quantization-aware training (\gls{QAT}) \cite{esser2019learned,he2023efficientdm} integrates the quantization process into the training or fine-tuning process of the model. During the forward pass, the weights are approximated using the chosen low-precision format. However, the rounding function is not differentiable. That is why in the backward pass a straight-through estimator (\gls{STE}) \cite{bengio2013estimating} is employed to approximate the gradients. Although the quantized weights are simulated during the forward pass, the high-precision weights are updated during training \cite{jacob2018quantization,courbariaux2015binaryconnect}. Upon completion of the training, the high-precision weights are permanently replaced with the quantized weights. While \gls{QAT} generally leads to superior performance compared to \gls{PTQ}, the full retraining of the model makes it computationally demanding \cite{gholami2022survey}.

The application of quantization to diffusion and flow-matching models presents unique challenges due to the iterative denoising process, wherein activations are highly time-dependent. Moreover, even minor quantization errors can exert a significant influence because of their accumulation over several inference steps. Consequently, diffusion-specific quantization methods have been developed to address these distinct challenges. Several approaches, such as post-training quantization for diffusion models (\gls{PTQD}) \cite{PTQD} and Q-Diffusion \cite{QDiffusion}, have been proposed. These methods leverage a timestep-aware calibration dataset by sampling features across the entire reverse diffusion process to compute optimal scaling factors. Building on this foundation, the authors of \cite{QDiT} successfully transferred these advanced quantization methods to \gls{DiT}s. A critical challenge inherited from Transformer-based architectures is the emergence of extreme activation outliers which is addressed by \cite{li2024svdquant}. They absorb these anomalous values into high-precision, low-rank components successfully compressing models like FLUX.1 [dev] to 4-bit precision. Most recently, \cite{FluxBit} demonstrated that the weights of the FLUX.1 [dev] can be compressed even further to 1.58-bit. 


Ultimately, quantization serves as an orthogonal technique to structural model pruning. As demonstrated by \cite{qu2025automatic}, both approaches can be successfully combined to achieve a further reduction in memory footprint.

\subsubsection{Non-Architectural VRAM Reduction}
Alongside methods that reduce the model parameter count, several strategies have been developed to minimize VRAM consumption during inference. 

An integral part of modern model architectures is the attention mechanism, which scales quadratically with the token sequence length, leading to a severe memory bottleneck. Consequently, hardware-aware algorithms such as FlashAttention \cite{FlashAttention} or xFormers \cite{lefaudeux2022xformers} were introduced to optimize memory access patterns and realize more efficient computation for large attention matrices. 

Furthermore, modern models such as FLUX.1 [dev] are typically deployed using the 16-bit half-precision format (FP16). A primary limitation of FP16 is its restricted dynamic range, meaning that gradient or activation outliers can easily exceed the maximum representable value, leading to numerical overflow that crashes both inference and training. For this reason, modern GPUs (e.g. NVIDIA A100, H100, and H200) natively support the BFloat16 format (BF16). BF16 utilizes the same 16-bit memory footprint as FP16 but preserves the dynamic range of FP32 by sacrificing fractional precision \cite{BF16}. Fortunately, neural networks have proven to be highly robust against this slight precision loss. 

 A third option for reducing \gls{VRAM} consumption is CPU offloading in which only specific subsets of the model are loaded onto the GPU, while currently inactive layers remain in the CPU memory.
  This results in slower inference and training, as an overhead is incurred when moving the model components between the GPU and CPU. Moreover, during training, optimizer states and gradients can also be offloaded to the CPU, further reducing the memory footprint. These memory management strategies are natively supported by modern training frameworks from Huggingface \cite{accelerate} or DeepSpeed \cite{ren2021zero,rajbhandari2021zero}. 
 
 \subsubsection{Model Pruning}
 Model pruning \cite{liu2020topological,janowsky1989pruning,mozer1989using} focuses on reducing the actual parameter count of models. It can be dissected into mainly two categories: unstructured \cite{nordstrom2022unstructured,frantar2023sparsegpt,han2015learning} and structured \cite{wen2016learning,anwar2017structured} model pruning. 
 
 In unstructured model pruning, individual, less important weights across the model's architecture are masked out so that they do not contribute to the inference process anymore. The process to identify parameters to prune can be based on several criteria, such as random weight selection \cite{liu2022unreasonable} or magnitude-based pruning \cite{han2015deep,frankle2018lottery}. Although this approach allows for the removal of many parameters while maintaining high performance and theoretically reducing the number of floating-point operations (FLOPs), it does not necessarily speed up execution or reduce \gls{VRAM} consumption in practice \cite{wen2016learning,gale2020sparse,ma2018shufflenet}. Simply setting parameters to zero does not alter the physical dimensions of the dense weight matrices, meaning the memory required to store them remains unchanged. To achieve an actual reduction in \gls{VRAM} consumption, these matrices need to be converted into a sparse tensor format, e.g. compressed sparse row (CSR) \cite{bell2008efficient}. However, because CSR stores the non-zero entries and their corresponding coordinates, it introduces an additional memory overhead that contradicts the efforts of \gls{VRAM} reduction \cite{hoefler2021sparsity}. Therefore, a high degree of sparsity is required to compensate for this overhead and ensure low memory requirements \cite{wen2016learning}. Moreover, modern GPUs are highly optimized for dense matrix multiplication, which is why running inference with sparse tensors on standard hardware does not necessarily result in a speed-up compared to their dense counterparts. Nevertheless, there are emerging custom kernels and specialized hardware to accelerate unstructured sparsity calculations \cite{gale2020sparse}. 
\\\\
 This Master's thesis focuses on the second pruning technique, namely structured pruning for Transformer-based diffusion and flow models. Unlike unstructured pruning, this method targets coherent model components, such as attention heads, channels, or entire Transformer blocks. To recover degraded performance and lost generation capabilities, most methods subsequently apply additional fine-tuning.
 Based on the compression strategy, the literature can be divided into two main categories, namely approaches that remove model components entirely (e.g. \cite{Bksdm, flux1-lite}) and those that replace them with a lightweight module (e.g. \cite{FastFlux,PPCL}).  
 
 The first category, encompassing complete component removal, has been widely applied in most research on compressing U-Net-based diffusion models, such as SDXL \cite{SDXL}. Despite this shared approach, these methods differ significantly in pruning granularity \cite{Koala,Laptop,Bksdm,ProgressiveDistillation,tang2023lightweight}. For instance, BK-SDM \cite{Bksdm} predominantly removed entire middle Transformer blocks from the SDXL architecture, identifying them as highly redundant. In contrast, LAPTOP-Diff \cite{Laptop} focused on pruning individual Transformer layers within the blocks, whereas KOALA \cite{Koala} employed a hybrid strategy that combined block removal with layer-wise pruning for higher compression ratios. Notably, KOALA observed that the decoder plays a more critical role than the encoder in the generation process and should therefore be subject to less aggressive compression.
 To recover the model's fidelity after pruning, most of these approaches leverage knowledge distillation \cite{KnowledgeDistillation} to fine-tune their pruned models, employing the unpruned model as a teacher to guide the recovery process. Addressing optimization challenges in this phase, LAPTOP-Diff \cite{Laptop} demonstrated that using normalized feature loss terms proved advantageous, as the magnitude of individual features from different layers may vary considerably. Furthermore, Progressive Distillation \cite{ProgressiveDistillation} showed that an iterative pruning schedule for SDXL leads to superior results compared to one-shot pruning. 
 
 Recently, the landscape of diffusion models has shifted from U-Net to purely Transformer-based architectures, where compression methods have become increasingly critical due to the steep increase in model size. The primary advantage of these architectures is their dimensional homogeneity. Unlike U-Net architectures, which may require down- or upsampling and channel adjustments, entire Transformer blocks can be removed without adjusting the feature dimension. 
 
 For this reason, \gls{DiT}s are ideal candidates for the first category of pruning, which involves the complete removal of Transformer blocks. Recent works, such as Flux.1 Lite \cite{flux1-lite} and TinyFusion \cite{TinyFusion}, have applied exactly this strategy. Flux.1 Lite compressed FLUX.1 [dev] to 8.1B parameters, while TinyFusion introduced a general compression framework for \gls{DiT}s that learns the blocks' task importance and prunes the most expendable ones. A third approach, HierarchicalPrune \cite{HierarchicalPrune}, is also based on removing entire Transformer blocks and subsequently applying an importance-aware training paradigm. 
 
 The second category involves replacing model components with lightweight modules. PPCL \cite{PPCL} is a hybrid approach leveraging a depth- and width-pruning paradigm. To reduce depth, the authors removed entire Transformer blocks. To complement this with width reduction, they identified that particularly the text stream and the \gls{MLP}s of the image stream in \gls{MMDiT}s exhibit high redundancy and replaced some of them with lightweight linear layers. Similarly, FastFLUX \cite{FastFlux} proposed replacing the residual branch of a ResNet block with a linear layer. To mitigate the performance loss \gls{LoRA}s were trained on the preceding and subsequent blocks to circumvent the computationally intensive fine-tuning of the entire model.
 
 In a distinct approach, Dense2MoE \cite{Dense2MoE} leverages a mixture-of-experts (\gls{MoE}) framework to dynamically activate only a small fraction of the model's parameters during each inference step. The core advantage of this approach is that it exploits the temporal variability of feature importance, recognizing that different Transformer blocks are critical at different timesteps. Their routing mechanism adaptively selects the most important blocks and experts for the current time step, making the model more expressive than standard static pruning methods. Additionally, the \gls{MLP}s within the blocks are replaced by smaller experts to further reduce the number of activated parameters. However, while the reduction in activated parameters per inference step significantly accelerates the generation process, it does not necessarily alleviate the \gls{VRAM} consumption bottleneck. Crucially, since nearly the entire model (including both activated and inactivated parameters) typically remains in memory, the total memory footprint stays large, which distinguishes this method significantly from memory-focused pruning methods. 
 
 Whereas all previous methods require subsequent fine-tuning, EcoDiff \cite{EcoDiff} provides a training-free approach by learning masks to achieve up to 20\% sparsity in FLUX.1 [dev].
 
 While all these methods focus on removing or replacing model parts of FLUX.1 [dev], a distinct compression category that remains largely unexplored for \gls{DiT}s is low-rank approximation, commonly implemented via truncated singular value decomposition (\gls{SVD}) 
 \cite{zhang2015singular}. In the domain of \gls{LLM} compression, \gls{SVD} has been extensively investigated as a post-training compression strategy \cite{DLowRankEst,SVD_LLM_V2}. Nevertheless, the straightforward application of truncated \gls{SVD} has proven to lead to significant performance drops \cite{hsu2022language,wang2025dobi,SVD_LLM_V2,wang2024svd} even at low compression ratios, as the singular value selection is solely based on weight magnitude and fails to account for the task importance \cite{GRASP}. 
To address this gap, this thesis revisits truncated \gls{SVD} within the context of \gls{DiT}s. Unlike in \gls{LLM}s, truncated \gls{SVD} is identified as a highly effective approximation strategy for projection matrices within Transformer blocks. Consequently, this work introduces \gls{SVD}trunc, an importance-aware rank-adaptive progressive compression schedule whose effectiveness is demonstrated on FLUX.1 [dev], establishing a new quality-parameter frontier. 


%This work demonstrates that an importance-aware rank-adaptive progressive compression scheme utilizing truncated \gls{SVD}, which is termed \gls{SVD}trunc, can effectively compress FLUX.1 [dev], outperforming current state-of-the-art pruning approaches on the HPSv2 \cite{HPSv2}, GenEval \cite{ghosh2023geneval} and DPG \cite{DPG} benchmarks. In addition, it is demonstrated that \gls{SVD} can be effectively deployed as a training-free compression approach for moderate compression ratios while maintaining acceptable performance. The thesis concludes with a proof-of-concept that the advanced \gls{SVD}-based compression method GRASP \cite{GRASP} can be transferred from \gls{LLM}s to flow models like FLUX.1 [dev].
 

 