\begin{titlepage}
\begin{center}
 
\Large\textbf{Department of Physics and Astronomy\\
Heidelberg University}

\vspace{18cm}

\normalsize
Master's Thesis in Physics\\
submitted by\\
\vspace{0.5cm}
\Large\textbf{Sebastian Heid}\\
\normalsize
\vspace{0.5cm}
born in Gelnhausen (Germany)\\
\vspace{0.5cm}
\Large\textbf{2026}
\normalsize


\newpage
\thispagestyle{empty}
\Large\textbf{Towards Efficient Diffusion Transformers: An Importance-Aware Low-Rank Compression Approach}
\vspace{18cm}

\normalsize
This Master's Thesis has been carried out by Sebastian Heid at the\\
Computer Vision and Learning Lab in Heidelberg\\
under the supervision of\\
\textbf{Prof. Dr. Carsten Rother} \\
and \\
\textbf{Prof. Dr. Tristan Bereau}
\vfill
\end{center}
\end{titlepage}

\thispagestyle{plain}
\pagenumbering{Roman}
\setcounter{page}{3}
\begin{center}
    \textbf{Abstract}\\
\end{center}
    Diffusion Transformers (\gls{DiT}s) are state-of-the-art (\gls{SOTA}) text-to-image generation models, capable of generating highly realistic and aesthetic images with exceptional detail. However, their immense size hinders their deployment in resource-constrained environments. While structural pruning reduces the parameter count of \gls{DiT}s, removing entire components severely disrupts the information flow, resulting in a clear performance drop even under moderate compression. Truncated singular value decomposition (\gls{SVD}) provides an alternative approach by approximating components rather than removing them entirely. Nevertheless, truncated \gls{SVD} has shown drastic performance degradation in large language models (\gls{LLM}s). In this work, truncated \gls{SVD} is revisited in the context of \gls{DiT}s, demonstrating its effectiveness. To investigate an \gls{SVD}-based compression pipeline, several preliminary experiments were conducted on PixArt-$\Sigma$, as its relatively low parameter count allows for extensive ablations. These findings culminated in the development of \gls{SVD}trunc, a novel compression protocol that iteratively selects the Transformer blocks based on their task importance and compresses less critical ones more aggressively, with subsequent fine-tuning utilizing knowledge distillation. Applying \gls{SVD}trunc to FLUX.1 [dev] and compressing it to 68\% of its original size results in only negligible loss in generation quality and outperforms all competing methods in terms of average performance across three distinct benchmarks (HPSv2, GenEval, and DPG). Further compression degrades performance smoothly even under substantial parameter reduction, which establishes a new quality-parameter frontier for FLUX.1 [dev]. Notably, \gls{SVD}trunc achieves strong performance even without additional fine-tuning, validating the robustness of low-rank approximation in \gls{DiT}s. Additionally, it generalizes to step-distilled \gls{DiT}s, which enables the development of extremely fast and memory-efficient models.
    

    
    \newpage
    \thispagestyle{plain}
    \begin{center}
    	\textbf{Zusammenfassung}\\
    \end{center}
    Diffusion Transformers (\gls{DiT}s) sind aktuell die leistungsstärksten Text-zu-Bild-Generatoren, welche realistische und ästhetisch ansprechende Bilder mit außergewöhnlicher Detailgenauigkeit erzeugen können. Ihre enorme Größe behindert jedoch ihren Einsatz in ressourcenbeschränkten Umgebungen. 
    Zwar reduziert strukturelles Pruning die Modellgröße, doch führt dies zu einer erheblichen Störung des Informationsflusses, was selbst bei moderater Komprimierung einen deutlichen Leistungsabfall zur Folge hat. Die trunkierte Singulärwertzerlegung (SVD) bietet einen alternativen Ansatz, indem sie Komponenten approximiert, anstatt sie vollständig zu entfernen. Dennoch hat die trunkierte SVD in großen Sprachmodellen (\gls{LLM}s) zu erheblichen Leistungsverlusten geführt.
     In dieser Arbeit wird die trunkierte SVD im Kontext von \gls{DiT}s erneut untersucht, wobei ihre Wirksamkeit für diese demonstriert wird. Zunächst wurden Experimente mit PixArt-$\Sigma$ durchgeführt, da dessen geringe Größe schnellere und umfangreichere Experimente zulässt. Die dadurch gewonnenen Erkenntnisse mündeten in die Entwicklung des Komprimierungsprotokolls SVDtrunc, das iterativ Transformer-Blöcke auf der Grundlage ihrer Aufgabenwichtigkeit auswählt und weniger kritische Blöcke stärker komprimiert. Anschließend findet ein Fine-Tuning des komprimierten Modells mittels Wissensdestillation statt. Die Anwendung von \gls{SVD}trunc auf FLUX.1 [dev] und die Komprimierung auf 68\% der ursprünglichen Größe führen zu einem vernachlässigbaren Verlust an Bildqualität und übertreffen alle konkurrierenden Methoden hinsichtlich der durchschnittlichen Leistung bei drei verschiedenen Benchmarks (HPSv2, GenEval und DPG). Außerdem zeigte sich, dass die Qualität mit weiterführender Kompression graduell und kontinuierlich abnimmt, was eine neue Qualitäts-Parametergrenze etabliert. Bemerkenswert ist, dass \gls{SVD}trunc auch ohne zusätzliches Fine-Tuning eine starke Leistung erzielt. Dies bestätigt die Robustheit der Low-Rank-Approximation in \gls{DiT}s. Darüber hinaus lässt es sich auf schritt-destillierte \gls{DiT}s verallgemeinern, was die Entwicklung extrem schneller und speichereffizienter Modelle ermöglicht.
     
     
     
    
    
%         Mit diesem  Kompressionsschema wurde FLUX.1 [dev] auf 68\% seiner Originalgröße verkleinert, wobei nur ein vernachlässigbarer Qualitätsverlust auftrat und \gls{SVD}trunc im Durchschnitt über alle drei Benchmarks (HPSv2, GenEval, DPG) den geringsten Leistungsabfall im Vergleich zu konkurrierenden Methoden aufwies
%    
    
%       Diffusion Transformers (\gls{DiT}s) are currently state-of-the-art (\gls{SOTA}) text-to-image generation models, capable of generating highly realistic and aesthetic images with exceptional detail. Despite these capabilities, their immense size makes them difficult to deploy in resource-constrained environments. Several structural pruning methods have been proposed to reduce the parameter count of \gls{DiT}s. Yet, the removal of individual components from the \gls{DiT} architecture severely disrupts the signal propagation through the network resulting in a clear performance drop even at moderate parameter reductions. Truncated singular value decomposition (\gls{SVD}) provides an alternative approach by approximating components rather than removing them entirely. However, truncated \gls{SVD} has shown drastic performance degradation in large language models (\gls{LLM}s) even under moderate compression. In this work, truncated \gls{SVD} is revisited in the context of \gls{DiT}s, and its effectiveness is demonstrated in contrast to previous findings in \gls{LLM}s. To investigate an \gls{SVD}-based compression pipeline for \gls{DiT}s, several preliminary experiments were conducted on PixArt-$\Sigma$, as its relatively low parameter count allows for extensive ablations. These findings culminated in \gls{SVD}trunc, a new compression protocol leveraging truncated \gls{SVD}. It iteratively selects the Transformer blocks based on their task importance and compresses less critical ones more aggressively, followed by subsequent fine-tuning utilizing knowledge distillation. \gls{SVD}trunc was applied to FLUX.1 [dev], an 11.9 billion parameter model, compressing it to 68\% of its original size. The compressed model demonstrates near-original generation quality and outperforms all competitive methods across three benchmarks (HPSv2, GenEval, and DPG). Further compression shows that the model performance degrades gradually and predictably even under substantial parameter reduction, establishing a new quality-parameter frontier for FLUX.1 [dev]. Furthermore, \gls{SVD}trunc achieves strong performance even without additional fine-tuning demonstrating the effectiveness of low-rank approximation in maintaining generation capabilities. In addition, it generalizes to step-distilled \gls{DiT}s positioning it as an complementary, orthogonal approach to existing inference step reduction techniques.
%    
